Dark Control: The Default Mode Network as a Reinforcement Learning Agent

### 2018-08-08 paper/paper.tex
edits to the stars

\begin{abstract}
The default mode network (DMN) is believed to subserve the
baseline mental activity in humans.
%
Its highest energy consumption compared to other brain networks and
its intimate coupling with conscious awareness are both pointing to
an overarching function.
%
Many research streams speak in favor of an evolutionarily adaptive role in
envisioning experience to anticipate the future.
In the present work, we propose a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
% DMN function is recast in terms of
% control theory and reinforcement learning by basing it on Markov decision processes.
%
The main purpose of the DMN, we argue,
may be to perform optimization based on Markov Decision Processes
through vicarious trial and error.
%
Our formal account of DMN function naturally accommodates as special cases
previous interpretations based on
(1) predictive coding,
(2) semantic associations, and
(3) a sentinel role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

This \textit{dark matter of brain physiology}
\citep{raichle2006brain} begs the question
of the biological purpose underlying DMN activity,
which however still remains elusive at the electrophysiological level
\citep{de2010temporal,brookes2011investigating,baker2014fast}.
% the link between the fMRI BOLD signal and population-level neural activity is still unclear. If those frequency-specific electrophysiological correlations are proposed as complementary to those observed with BOLD \citep{hipp2015bold}, their role in DMN function remains elusive \citep{maldjian2014graph}.
What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
%
Axonal tracing injection in such parts of the association cortex in monkeys
were shown to resemble connectivity links between nodes of the human
DMN (see here for details on anatomical connections: \citep{randy2008}).
Additionally, myelination patterns of axon connections were found to finish
particularly late in these cortical areas \citep{flechsig1920}, often believed to suggest
sophistication of subserved neural processes
\citep{Sowell2003, Yakovlev1967}.
%
We propose that this set of some of the most advanced regions in
the association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The network nodes that compose the human DMN
are responsible for extended parts of the baseline neural activity,
which typically decreases when engaged in controlled psychological experiments
\citep{raichle_baseline}.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict the environment using
mental imagery as an evolutionary advantage \citep{suddendorf2007evolution}.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this brain network appears to
simultaneously modulate perception-action cycles in the present
and to support mental travel
across time, space, and content domains
\citep{boyer2008evolutionary}.

The present work adopts the perspective of
a human \textit{agent} faced with the choice of the next actions
and guided by outcomes to optimize behavioral performance. These outcomes
can be really experienced, hypothetically imagined, or simply expected in the future.
Formally, we propose reinforcement learning as a particularly attractive framework
for describing, containing, and quantifying the DMN function.
An intelligent agent improves the interaction with the environment
by continuously updating its computation of value estimates and action predispositions
through integration of feedback outcomes.
That is, ``[agents], with their actions, modify the environment and in doing so
partially determine their next stimuli, in particular stimuli that are necessary
for triggering the next action''
\citep{pezzulo2011grounding}.
Agents with other behavioral policies therefore sample different
distributions of action-perception trajectories \citep{ghavamzadeh2015bayesian}.
Henceforth,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to reach preferred states.

@@

Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.
We argue that DMN implication in many
of the most advanced human capacities
can be recast as prediction error minimization
informed by internally generated probabilistic simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
allowing maximization of action outcomes across multiple time-scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Monte Carlo sampling.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide randomly sampled blocks of pseudo-experience instrumental to iteratively
optimize the behavior of the organism.

Evidence from computational modeling of human behavior \citep{kording2004bayesian} and cell recording experiments in ferrets \citep{fiser2004small} suggest that
the brain is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The present paper proposes a
process model that satisfies this previously proposed contention.
We also contribute to the discussion of DMN function
by providing tentative evidence that
variation of the grey-matter volume in DMN regions
is linked to the reward circuitry (Fig. 2),
thus linking two literatures with currently scarce cross-references.
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments.
At this stage, we emphasize the importance of differentiating which levels of observation are at play in the present account. A process model is not solely intended to capture behavior of the agent, such as cognitive accounts of DMN function, but also the neurocomputational specifics of the agent. Henceforth, we will use ``inference`` when describing aspects of the statistical model, ``prediction`` when referring to the neurobiological implementation, and words like ``forecast`` or ``forsee`` when referring to the behavior of the agent.
%
Notably, our account does not claim that the brain or neural activity in the DMN are completely identical with reinforcement learning algorithms. Rather, we advocate feedback-based learning strategies as an attractive alternative perspective
to understand DMN function.

@@

\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
    &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{\operatorname{{KL}}(p_R(\z|\x) \| p_G(\z|\x))}_{\text{complexity}} \\
    &\le F^R_G(\x),
    \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z.
  \end{split}
\end{equation}
where
\begin{equation}
  F_G^R(\x) := \langle -\log(p_G(\z, \x))\rangle_{p_R(\z|\x)} - \mathcal H(p_R(\z|\x))
\end{equation}
is the \textit{free energy}.
Here, the angular brackets denote the \textit{expectation} of the joint negative log-likelihood $-\log(p_G(\z, \x))$ w.r.t the recognition density $p_R(\z|\x)$, $\mathcal H$ is the \textit{entropy} functional defined by $\mathcal H(p) := -\sum_{\z}p(\z)\log(p(\z)$, while $\operatorname{KL}(.\|.)$ is the usual \textit{Kullback-Leibler (KL) divergence} (also known as \textit{relative entropy}) defined by $\operatorname{KL}(p\|q) := \sum_{z}p(\z)\log(p(\z)/q(\z)) \ge 0$,
which is a measure of how different two probability distributions are. In this framework, the goal of the agent is then to iteratively refine the generative model $p_G$ and the recognition model $p_R$ so as to minimize the free energy $F_G^R(\x)$ over sensory input $\x$.

Importantly, $F^R_G(x)$ gets low in the following cases:
\begin{itemize}
\item  $p_R(\z|\x)$ puts a lot of mass on configurations $(\z,\x)$ which are $p_G$-likely, % (i.e the sensation $\x$ is likely caused by $\z$)
  \textbf{and}
\item $p_R(\z|\x)$ is as uniform as possible (i.e have high entropy), so as not to concentrate
  all its mass on a small subset of possible causes for the sensation $\x$.
  \end{itemize}
% The optimization procedure proposed by the authors derives from the \textit{wake-sleep algorithm} \citep{dayan1995helmholtz}.
% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}

% CRITICISM
% \paragraph{Main criticisms.}
Despite its popularity, criticism against the FEP has been voiced over the years,
some of which is outlined in the following.
The main algorithm for minimizing free energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As these authors noted,
a crucial drawback of the wake-sleep algorithm (and therefore of theories like the FEP \citep{friston2010free}) is that it involves a pair
of forward (generation) and backward (recognition) models $p_G$ and $p_R$ that
together does not correspond to optimization of (a bound of) the marginal likelihood
because KL divergence is not symmetric in its arguments.

@@

would remain trapped in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
% \begin{wrapfigure}{r}{.37\linewidth}
%   \centering
%   \includegraphics[width=1.\linewidth]{darkroom_resized.png}
%   \caption{\textbf{The dark room experiment.} An intelligent agent situated
%   in a light-deprived closed space is used
%   as a thought experiment for the complete absence
%   of external sensory input.
%   % The free-energy principle theory paradoxally predicts that such an agent would remain quietly in this room, such it perfectly predict all sensory input from such an environment.
%   % However such behavior is not observed in humans or any animals: the agent would strive to leave such an environment in search of more ``interesting'' worlds.
%   }
%   \label{fig:darkroom}
% \end{wrapfigure}
However, such a behavior is seldom observed in humans in the
real world. In a dark room, the intelligent agents would search for light sources by explore
the surroundings or aim to exit the room.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high relevance.
Driven by the surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would inevitably
identify the sensory-stimulus-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) could serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or finding the room exit.

% \paragraph{Links between FEP and RL.}
We finally note that FEP and active inference can be reframed in terms of
our RL framework.
This becomes possible by recasting the Q-value function (i.e., expected long-term reward) maximized by the DMN to
correspond to negative surprise,
that is, the log-likehood of current sensory priors
the agent has about the world. More explicitly, this corresponds to using free-energy as a Q-value approximator for the MDP such that

>>>>

\begin{abstract}
The default mode network (DMN) is believed to subserve the
baseline mental activity in humans.
%
Its highest energy consumption compared to other brain networks and
its intimate coupling with conscious awareness are both pointing to
an overarching function.
%
Many research streams speak in favor of an evolutionarily adaptive role in
envisioning experience to anticipate the future.
In the present work, we propose a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
% DMN function is recast in terms of
% control theory and reinforcement learning by basing it on Markov decision processes.
%
The main purpose of DMN activity, we argue,
may be to optimize action policies via value estimates
based on Markov Decision Processes through vicarious trial and error.
%
Our formal perspective on DMN function naturally accommodates as special cases
previous interpretations based on
(1) predictive coding,
(2) semantic associations, and
(3) a sentinel role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

This \textit{dark matter of brain physiology}
\citep{raichle2006brain} begs the question
of the biological purpose underlying neural activity in the DMN,
which however still remains elusive at the electrophysiological level
\citep{de2010temporal,brookes2011investigating,baker2014fast}.
% the link between the fMRI BOLD signal and population-level neural activity is still unclear. If those frequency-specific electrophysiological correlations are proposed as complementary to those observed with BOLD \citep{hipp2015bold}, their role in DMN function remains elusive \citep{maldjian2014graph}.
What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
%
Axonal tracing injection in such parts of the association cortex in monkeys
were shown to resemble connectivity links between nodes of the human
DMN (see here for details on anatomical connections: \citep{randy2008}).
Additionally, myelination patterns of axon connections were found to finish
particularly late in these cortical areas \citep{flechsig1920}, often believed to suggest
sophistication of subserved neural processes
\citep{Sowell2003, Yakovlev1967}.
%
We propose that this set of some of the most advanced regions in
the association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The network nodes that compose the human DMN
are hubs of high baseline neural activity,
which typically decreases when engaged in controlled psychological experiments
\citep{raichle_baseline}.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict the environment using
mental imagery as an evolutionary advantage \citep{suddendorf2007evolution}.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this brain network appears to
simultaneously modulate perception-action cycles in the present
and to support mental travel
across time, space, and content domains
\citep{boyer2008evolutionary}.

We aim at proposing an alternative to reasoning about the DMN
based on longstanding cognitive theory.
The present work adopts the control-theoretical perspective of
a human \textit{agent} faced with the choice of the next actions
and guided by outcomes to optimize behavioral performance. These outcomes
can be really experienced, hypothetically imagined, or simply expected in the future.
Formally, we propose reinforcement learning as a particularly attractive framework
for describing, containing, and quantifying the unknown function underlying DMN activity.
An intelligent agent improves the interaction with the environment
by continuously updating its computation of value estimates and action predispositions
through integration of feedback outcomes.
That is, ``[agents], with their actions, modify the environment and in doing so
partially determine their next stimuli, in particular stimuli that are necessary
for triggering the next action''
\citep{pezzulo2011grounding}.
Agents with other behavioral policies therefore sample different
distributions of action-perception trajectories \citep{ghavamzadeh2015bayesian}.
Henceforth,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to reach preferred states.

@@

Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.
We argue that DMN implication in many
of the most advanced human capacities
can be recast as prediction error minimization
informed by internally generated probabilistic simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
allowing maximization of action outcomes across different time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Monte Carlo sampling.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide randomly sampled blocks of pseudo-experience  that are
instrumental to iteratively
optimize the behavioral agenda of the organism.

Evidence from computational modeling of human behavior \citep{kording2004bayesian} and cell recording experiments in ferrets \citep{fiser2004small} suggest that
much of brain activity is dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The present paper proposes a
process model that satisfies this previously proposed contention.
We also contribute to the discussion of DMN function
by providing tentative evidence that
variation of the grey-matter volume in DMN regions
is linked to the reward circuitry (Fig. 2),
thus linking two literatures with currently scarce cross-references.
Finally,
we derive explicit hypotheses that could be tested in targeted neuroscience
experiments in the future
and
we detail how our process model relates
to previous cognitive and theoretical accounts of DMN function.

Please appreciate the importance of differentiating which levels of observation are at play in the present account. A process model is not solely intended to capture behavior of the agent, such as cognitive accounts of DMN function, but also the neurocomputational specifics of the agent. Henceforth, we will use ``inference`` when referring to aspects of the statistical model, ``prediction`` when referring to the neurobiological implementation, and words like ``forecast`` or ``forsee`` when referring to the behavior of the agent.
%
It is moreover important to note that
our account does not claim that neural activity in the DMN in particular
or the brain in general are identical with reinforcement learning algorithms.
Rather, we advocate feedback-based learning strategies as an attractive alternative perspective
to describe, quantify, and interpret DMN function in neuroscience research.

@@

\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
    &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{\operatorname{{KL}}(p_R(\z|\x) \| p_G(\z|\x))}_{\text{complexity}} \\
    &\le F^R_G(\x),
    \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z.
  \end{split}
\end{equation}
where
\begin{equation}
  F_G^R(\x) := \langle -\log(p_G(\z, \x))\rangle_{p_R(\z|\x)} - \mathcal H(p_R(\z|\x))
\end{equation}
is the \textit{free energy}.
Here, the angular brackets denote the \textit{expectation} of the joint negative log-likelihood $-\log(p_G(\z, \x))$ w.r.t the recognition density $p_R(\z|\x)$, $\mathcal H$ is the \textit{entropy} function defined by $\mathcal H(p) := -\sum_{\z}p(\z)\log(p(\z)$, while $\operatorname{KL}(.\|.)$ is the usual \textit{Kullback-Leibler (KL) divergence} (also known as \textit{relative entropy}) defined by $\operatorname{KL}(p\|q) := \sum_{z}p(\z)\log(p(\z)/q(\z)) \ge 0$,
which is a measure of difference between two probability distributions. In this framework, the goal of the agent is to iteratively refine the generative model $p_G$ and the recognition model $p_R$ so as to minimize the free energy $F_G^R(\x)$ over sensory input $\x$.

Importantly, $F^R_G(x)$ gets low in the following cases:
\begin{itemize}
\item  $p_R(\z|\x)$ puts a lot of mass on configurations $(\z,\x)$ which are $p_G$-likely
% (i.e the sensation $\x$ is likely caused by $\z$)
%  \textbf{and}
\item $p_R(\z|\x)$ is as uniform as possible (i.e have high entropy), so as not to concentrate
  all its mass on a small subset of possible causes for the sensation $\x$
  \end{itemize}
% The optimization procedure proposed by the authors derives from the \textit{wake-sleep algorithm} \citep{dayan1995helmholtz}.
% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}

% CRITICISM
% \paragraph{Main criticisms.}
Despite its popularity, criticism against the FEP has been voiced over the years,
which we allude to in the following.
The main algorithm for minimizing free energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As these authors noted,
a crucial drawback of the wake-sleep algorithm (and therefore of theories like the FEP \citep{friston2010free}) is that it involves a pair
of forward (generation) and backward (recognition) models $p_G$ and $p_R$ that
together does not correspond to optimization of (a bound of) the marginal likelihood
because KL divergence is not symmetric in its arguments.

@@

would remain trapped in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
% \begin{wrapfigure}{r}{.37\linewidth}
%   \centering
%   \includegraphics[width=1.\linewidth]{darkroom_resized.png}
%   \caption{\textbf{The dark room experiment.} An intelligent agent situated
%   in a light-deprived closed space is used
%   as a thought experiment for the complete absence
%   of external sensory input.
%   % The free-energy principle theory paradoxally predicts that such an agent would remain quietly in this room, such it perfectly predict all sensory input from such an environment.
%   % However such behavior is not observed in humans or any animals: the agent would strive to leave such an environment in search of more ``interesting'' worlds.
%   }
%   \label{fig:darkroom}
% \end{wrapfigure}
However, such a behavior is seldom observed in humans in the
real world. In a dark room, the intelligent agents would search for light sources by explore
the surroundings or aim to exit the room.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is framed as being multi-scale, and there would be one scale at which
the organism would
be \textit{surprised} by a \textit{lack of surprise}.
One may object that a dark room would paradoxically correspond to a state of particularly high relevance.
Driven by the surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would inevitably
identify the sensory-stimulus-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) could serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input,
potentially subserved by the DMN, could then entice the organism to explore the room,
for instance by looking for and using the light switch or finding the room exit.

% \paragraph{Links between FEP and RL.}
We finally note that FEP and active inference can be reframed in terms of
our RL framework.
This becomes possible by recasting the Q-value function (i.e., expected long-term reward) maximized by the DMN to
correspond to negative surprise,
that is, the log-likehood of current sensory priors
the agent has about the world. More explicitly, this corresponds to using free-energy as a Q-value approximator for the MDP in the following way:
